Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-17 latex/main.tex
Updates from Overleaf

\begin{itemize}
\item Snakemake workflow for aligning and converting the Viridian dataset to
VCF Zarr:
\url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/viridian\_dataset/}
\end{itemize}

TODO add these to the list above with actual URLs
\begin{itemize}
    \item The ARG and the sequence alignments have been deposited at Zenodo and are publicly available.
    \item All original code has been deposited in GitHub and is publicly available.
    \item Jupyter notebooks to perform the analyses in this paper are available on GitHub.
\end{itemize}

@@

% \subsection*{Key resources table}

% \textit{To create the KRT, please use the \href{https://star-methods.com}{KRT webform} or the \href{http://www.cell.com/pb-assets/journals/research/cell/methods/table-template1.docx}{Word template} and upload this file separately.}

@@

Sc2ts is implemented in Python building on core PyData infrastructure
such as Numpy\cite{Harris2020array}, SciPy\cite{Virtanen2020scipy}
and Pandas~\cite{Mckinney2010data} ,
using pyfaidx\cite{Shirley2015efficient} to access alignment data
and Numba~\cite{Lam2015numba} to accelerate certain alignment processing
tasks.
[TODO JK to fill this out]

@@

\subsubsection*{Pandemic scale alignment storage}
To provide the efficient access to alignment data required for inference with sc2ts, we
developed a new format based on the VCF Zarr~\cite{Czech2025} specification.
Briefly, Zarr is a data storage format designed for scientific data applications
originally developed to store genetic variation data for the
\textit{Anopheles gambiae}
1000 Genomes Project~\cite{Anopheles2017genetic}, and
is now seeing adoption across a wide range of 
applications~\cite{Abernathey2021cloud,Moore2021ome,Gowan2022using,
Dhapola2022scarf,Virshup2023scverse,Baker2023emobject,Marconato2024spatialdata,Ruan2024image}.
Zarr is essentially a simple mechanism for storing n-dimensional array
data as a regular grid of compressed chunks, which provides 
high levels of data compression, efficient access across multiple
dimensions, and is ideally suited to modern cloud-based
deployments~\cite{Czech2025}.

We converted the Viridian v05 dataset\cite{Hunt2024}, consisting of 4,484,157 SARS-CoV-2
consensus sequences (125GiB over 48 FASTA files) and 30 metadata fields (1.4
GiB TSV file) to VCF Zarr using a 
reproducible Snakemake\cite{Koster2012snakemake} pipeline.
We aligned each sequence to the Wuhan-Hu-1 reference sequence (MN908947.3) using 
MAFFT v7.475 \cite{Katoh2013},
with the option \texttt{--keeplength} to retain the reference coordinate system.
We then used \texttt{sc2ts import-alignments} and 
\texttt{sc2ts import-metadata} to convert all alignments and metadata to VCF
Zarr format, resulting in a single 401 MiB file 
(around 77X smaller than gzip compressed FASTA\cite{Czech2025}).

A classical tradeoff in bioinformatics is whether to store data 
in sample-major (FASTA) or variant-major (VCF) format, determining
whether the data can be accessed efficiently either by row or column
of the variant matrix. Because VCF Zarr is chunked in two dimensions,
we can now access either sample aligments or the variation
data for a given site across all samples efficiently.
See Czech et al.\cite{Czech2025} for more details and benchmarks on 
this dataset.  
Another advantage of Zarr over formats such as VCF is that is
is inherently ``appendable'', and large datasets can be updated
in real time.

Both the pipeline and converted Viridian dataset are available for download;
see Data and code availability for details.

\subsubsection*{Sample preprocessing}
The collection date for a sequence is an important piece of information
for sc2ts. We used the Viridian
consensus date (``Date\_tree``), based on information from COG-UK, GISAID, and
ENA/SRA as our source. Samples that have missing or incomplete dates (e.g., 2020-01)
were omitted, as were
the $\sim$40,000 samples which had a collection date of December 31, 2020.
As the  sampling frequency of this date is much higher than other
(at most 15,000 samples per day), it seems likely that this 
batch is highly enriched for wrong collection dates.
Furthermore, manual checking (by cross-referencing the ENA and COG-UK Mutation Explorer)
confirmed that some of the samples were not collected in 2020.

The bioinformatics community has identified and curated a set of problematic sites
for SARS-CoV-2 that are enriched for systematic sequencing errors or 
high levels of homoplasy~\cite{DeMaio2020issues}. 
As these sites had not been updated to
take into account the systematic improvements made by the Viridian dataset,
we assembled a custom set of sites to mask. 
To do so, we first built a preliminary ARG from the samples collected up to June 30, 2021,
without any site masking.
Then, we identified the top 100 sites which had the highest mutation counts in
the ARG.
Only 11 of these sites 
(635, 8835, 11074, 11083, 15521, 16887, 21304, 21305, 21575, 21987 and 28253)
were present in the problematic sites file
(\url{https://raw.githubusercontent.com/W-L/ProblematicSites\_SARS-CoV2/master/problematic\_sites\_sarsCov2.vcf}).

Alignments read from the VCZ Zarr file were first preprocessed to mask any 
non nucleotide characters (A,C,G or T) as missing data. 
Then, we filtered 
any samples that had missing data at more than 500 sites, excluding
the 100 sites identified as problematic (this was 
set to 10,000 sites for the period up to 2020-03-01).

@@

This computation is a good example of type of integrated large-scale data analysis 
that can be easily performed using the combination of tskit and VCF Zarr.
The entire analysis is performed in a short Jupyter
notebook~\cite{kluyver2016jupyter}, with the 
core calculation requiring about 30 lines of Python code. Retrieving 
data from the base alignments and ARGs for 2.48 million samples 
at the same time over all 27,507 sites took only 6 minutes, with a peak memory
usage of about 5GiB of RAM on a standard desktop computer.

>>>>

\begin{itemize}
    \item The ARG and the sequence alignments have been deposited at Zenodo and are publicly available.
    \item All original code has been deposited in GitHub and is publicly available.
    \item Jupyter notebooks to perform the analyses in this paper are available on GitHub.
\end{itemize}

@@

\subsection*{Key resources table}

\textit{To create the KRT, please use the \href{https://star-methods.com}{KRT webform} or the \href{http://www.cell.com/pb-assets/journals/research/cell/methods/table-template1.docx}{Word template} and upload this file separately.}

@@

Sc2ts is implemented in Python building on core PyData infrastructure
such as Numpy\cite{Harris2020array}, SciPy\cite{Virtanen2020scipy}
and Pandas ,
using pyfaidx\cite{Shirley2015efficient} to access alignment data
[TODO JK to fill this out]

@@

\subsubsection*{Data preprocessing}

In this study, we used the SARS-CoV-2 sequences reassembled in a recent study \cite{Hunt2024}
using an amplicon-aware assembly method (named ``Viridian'').
The Viridian sequences contain far fewer artifacts compared to their publicly deposited counterparts on GenBank.
Consequently, phylogenetic tree inference using the Viridian data led to more parsimonious phylogenies,
as seen by a large reduction in the number of likely artifactual reversions \cite{Hunt2024}.

We downloaded 4,484,157 Viridian sequences obtained
from samples collected from January, 2020 to March, 2023 (``batch 1'').
We aligned each sequence to the Wuhan-Hu-1 reference sequence (MN908947.3) using MAFFT v7.525 \cite{Katoh2013},
with the option ‘--keeplength’ to retain the reference coordinate system, discarding insertions.
We converted the MAFFT alignments and all sample metadata to VCF Zarr format \cite{Czech2025},
which are available for download on Zenodo (URL).

Prior to ARG inference, we preprocessed the remaining sequences as follows.
We treated low-quality bases (represented by Ns), gaps and ambiguous characters in the sequences as missing data.
When encoded as missing data, the bases do not inform sample matching (see ``The Li and Stephens model'').

During inference, we filtered out the samples with low-quality sequences and low-precision sampling dates.
We kept the quality check-passing sequences used to infer the Viridian UShER phylogeny \cite{Hunt2024}.
We skipped low-quality sequences that had more than 500 missing sites
(except when building the ``base'' ARG, when the maximum number of allowed missing bases was set to 10,000 instead).
We also skipped $\sim$40,000 samples which had a collection date of December 31, 2020.
The sampling frequency of this date is alarmingly high compared to other dates (at most 15,000 samples per day).
We suspect that this batch of samples might be unusually enriched for wrong collection dates,
likely due to arbitrary assignment of the originally reported collection date of ``2020'' to be ``31-12-2020''.
Furthermore, manual checking (by cross-referencing the ENA and COG-UK Mutation Explorer)
confirmed that some of the samples were not collected in 2020.

Sites which are enriched for systematic sequencing errors or frequently experience homoplasy
can cause artifactual signals of recombination (REF).
The bioinformatics community has identified and curated a set of problematic sites
that are recommended to be excluded before phylogenetic inference (REF).
Here we assembled a bespoke set of such problematic sites
using the information in the Viridian sequences.
To do so, we first built a preliminary ARG from the samples collected up to June 30, 2021,
without any site masking.
Then, we identified the top 100 sites which had the highest mutation counts in the ARG (URL).
Only six of these sites overlapped with the ``Problematic Sites''
downloaded from the UCSC Genome Browser website (URL),
which did not have a high number of mutations in the preliminary ARG.
To infer the final ARG, we masked the problematic sites
in our custom set
\url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian\_config.toml}.

@@

This computation is a good example of type of integrated large-scale data analysis 
that can be easily performed using the combination of tskit and VCF Zarr.
The entire analysis is performed in a short Jupyter notebook, with the 
core calculation requiring about 30 lines of Python code. Retrieving 
data from the base alignments and ARGs for 2.48 million samples 
at the same time over all 27,507 sites took only 6 minutes, with a peak memory
usage of about 5GiB of RAM on a standard desktop computer.
